Soss

### 2019-12-31 paper/paper.tex
example model
edits

There are many approaches to building a Probabilistic Programming Language (``PPL''). Soss is distinct from most in a number of ways...

\section{Model syntax}

\begin{figure}[H]
    \centering

    \begin{subfigure}[b]{0.4\textwidth}
        \begin{tikzpicture}

@@

        \end{tikzpicture}
        \caption{}
    \end{subfigure}
    \hspace{0.1in}
    \begin{subfigure}[b]{0.4\textwidth}

@@

    \caption{A plate diagram (a) and Soss encoding (b) for a two-way analysis of variance model. Note the specification that $y$ will later be observed is not part of the model, but is instead given at inference time.}
    \label{fig:model}
\end{figure}

Figure~\ref{fig:model} shows a plate diagram and Soss implementation of a simplified two-way ANOVA model. The model $m$ has \emph{arguments} $I$ and $J$. Each line of a model is either a \emph{sample statement} (\jl#v ~ rhs#) or an \emph{assign statement} (\jl#v = rhs#). In either case, \jl#v# must be a valid Julia variable name, and \jl#rhs# must be a valid Julia expression. For a sample statement, \jl#rhs# should also be ``distribution-like'', supporting \jl#logpdf# and/or \jl#rand# methods (depending on the inference algorithm to be used).

\section{Inference Primitives}

@@

\begin{figure}

    \begin{minted}{julia}
        julia> truth = rand(m(I=2,J=3)); pairs(truth)
        pairs(::NamedTuple) with 10 entries:
          :I    => 2
          :J    => 3
          :yhat => [0.0996722 0.720247 -0.706772; 0.100505 0.72108 -0.705939]
          :σ    => 0.297288
          :λ    => 0.523386
          :σJ   => 0.312794
          :σI   => 0.0054669
          :β    => [0.0973136, 0.717889, -0.70913]
          :α    => [0.00235854, 0.00319107]
          :y    => [0.386041 0.564963 -0.72177; 0.236898 0.842491 -0.912154]
    \end{minted}

\end{figure}

\section{Model Transformations}

\subsection{Conditional Predictive models}

\subsection{Causal Interventions (``Do'' operator)}

\subsection{Markov Blankets}

@@

\section{Symbolic Manipulation}

\subsection{Symbolic Log-density}

\subsection{Code Generation}

>>>>

There are many approaches to building a Probabilistic Programming Language (``PPL''). Soss is distinct from most alternative in a number of ways...

\begin{itemize}
    \item Model specification in Soss is \emph{declarative}; statements are ordered not by the order entered by the user, but by the partial order given by their variable dependencies.
    \item Soss models are \emph{first-class}; models can be passed as arguments to other models, or used in place of distributions within a model.
    \item Values and distributions in Soss are represented internally as \emph{abstract syntax trees}, maximizing the flexibility of available inference methods.
    \item Code generation \texttt{GeneralizedGenerated.jl}
    \item Model transformations
    \item Function-like models
    \item Symbolic simplification
    \item Extensibility
\end{itemize}

\section{Modeling in Soss}

Figure~\ref{fig:model} shows a plate diagram and Soss implementation of a simplified two-way ANOVA model. The Soss model $m$ has \emph{arguments} $I$ and $J$. Each line of a model is either a \emph{sample statement} (\jl#v ~ rhs#) or an \emph{assign statement} (\jl#v = rhs#). In either case, \jl#v# must be a valid Julia variable name, and \jl#rhs# must be a valid Julia expression. For a sample statement, \jl#rhs# should also be ``distribution-like'', supporting \jl#logpdf# and/or \jl#rand# methods (depending on the inference algorithm to be used).

\begin{figure}[t]
    \centering

    \begin{subfigure}[b]{0.3\textwidth}
        \[
    \begin{aligned}
        \\ i &\in \{1,\cdots,I\}
        \\ j &\in \{1,\cdots,J\}
        \\ λ &\sim \text{Cauchy}_+(0,1)
        \\ \sigma_I &\sim \text{Normal}_+(0,\lambda)
        \\ \sigma_J &\sim \text{Normal}_+(0,\lambda)
        \\ \alpha_i &\sim \text{Normal}(0,\sigma_I)
        \\ \beta_j &\sim \text{Normal}(0,\sigma_J)
        \\ \hat{y}_{ij} &= \alpha_i + \beta_j
        \\ y_{ij} &\sim \text{Normal}(\hat{y}_{ij}, \sigma)
    \end{aligned}
    \]

    \caption{}
    \end{subfigure}
    \begin{subfigure}[b]{0.3\textwidth}
        \scalebox{0.85}{\begin{tikzpicture}

@@

        \end{tikzpicture}}
        \caption{}
    \end{subfigure}
    \hspace{0.1in}
    \begin{subfigure}[b]{0.3\textwidth}

@@

    \caption{A generative model (a), plate diagram (b), and Soss model (c) for a two-way analysis of variance model. Note the specification that $y$ will later be observed is not part of the model, but is instead given at inference time.}
    \label{fig:model}
\end{figure}

\subsection{Inference Primitives}

@@

\subsection{Model Transformations}

\subsubsection*{Conditional predictive models}

\subsubsection*{Causal interventions (``Do'' operator)}

\subsubsection*{Markov blankets}

@@

\subsection{Symbolic Manipulation}

\subsubsection*{Symbolic Log-density}

\subsubsection*{Code Generation}
